Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-01 demystify.tex
updates

\paragraph{Interviews} Just like a job application, you likely will get interviews from schools. The most common case is that a professor is interested in working with you and wants to chat to make decision e.g., to offer RA (\autoref{sec:ra})). In some cases, the interview is done by several professors, e.g., to see if a student fits to their group or to recruit a very strong student to their program. The interview can also be done by adcom members to see if the student fit their program.
%The professor might include one of their current students as part of the interview.

@@

\textbf{Advice}: Treat the interview as an informal chat. Have elevator pitches about your research experiences and interests. You might also want have some 5-minute presentation about your research. If a professor asks you to read a paper, do it and be prepared to discuss it. You can also ask if you need to prepare for a coding problem. Finally, prepare some thoughtful questions to ask, e.g., about the program, the university, or the professor's research.

@@

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few  would have NSF Graduate or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even \emph{``honorable mentioning''}, school will still mention you on their website, and you should put it on your CV.

@@

\begin{commentbox}[Hung:]
  I always encourage the students I admitted to talk with my students and the students of other faculty in other schools who admitted them. You will unlikely hear straight-out complaints from current students in a professor’s group. But sometimes what is important are things that they (current students) don’t tell you. Pay attention to their ``level of excitement'' being in the group.
\end{commentbox}

@@

Here is some general advice for finding a professor.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research that professor and contact them (\autoref{sec:contact}).

@@

Another effective way is taking graduate-level courses in the topics you are interested in.  Professors teaching \emph{special topics courses} and \emph{research seminars} might be recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the prof after classes, etc---being stand out.  Many professors, including myself, prefer taking in new students this way.  It gives both the professor and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).

@@

\paragraph{Not a good fit}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if there is no faculty who is willing to advise you (e.g., they are on sabbatical, personal leave, do not have sufficient funding, or already have too many students already), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

For example, 2024 admission cycle has a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Consequently, AI/NLP faculty might be overwhelmed and cannot consider many candidates, even those with excellent profiles.

@@

\paragraph{Overqualified or Lack Interest}  This might be paradoxical and is the opposite of aiming too high. However if adcom believes that you are likely to get admitted and go to a better program (i.e., they are just your ``safety'' schools), they might not be so enthusiastic to admit you and want to save the spot for someone else. Note though it does happen, for most good schools this is rare and you should not assume this is the case.

@@

% \section{How much time do you spend looking at an application?}\label{sec:time-spent}

@@

% I first skim the summary for school, GPA, test scores and looking for red flags like low GPA or test scores that do not meet the minimum requirements. Then I scan transcripts for low/failing grades on CS and related courses.  For these I make comments such as ``- unknown international school with good GPA, too many low grades on CS and related courses" or ``+ very good GPA from top school XYZ". I do not look at test details as the summary already gives the main scores.

% Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experiences to distinguish between a strong LoRs from the rest). My comments for LoRs include ``+ very strong LoRs from well-known prof. XYZ talking about research experiences'' or ``- generic LoRs, mostly about course work''.  I also note any unusual or unique details in the LoRs, e.g., ``+ LoR mentioned about student's significant contribution on well-known open-source project X''. Sometimes a LoR came directly from a GMU professor who mentioned they already know the student and intend to offer RA, which is a very strong signal for admission (and my comment would be ``Prof. X making RA offer'').

% I skim quickly over CV, looking for paper publications, research work experiences, and other relevant achievements such as winning prizes in well-known international programming/math competitions.  My notes for CV include things such as ``+ has X papers at top venues'' or ``+ gold medal at X''. Sometimes I also note about unusual or unique experiences, e.g., ``+ has a patent on X''.

% Next is the SOP, which I read carefully (but if it is bad then it would a quick read). My notes for SOP include things such as ``- SOP mostly on course work, no clear motivation, generic" , or ``+ very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned about publications then I would have ``+ has X papers at top venues" or ``+/- has X papers but at unknown places''.  I will not go online search for the papers (unless they upload them as sample papers).  I also skim over sampling papers, essentially looking over the abstract to see if reads well, and make notes such as ``+ have a paper under submission to unknown venue, but reads well". If the student mentioned about a project, I might also look at it (e.g., if it is on Github) and make notes such as ``+ has project on Github with X stars''.

% At this point I have enough to make up my decision, which I enter to the evaluation system. Typically my decisions are either \emph{reject} or \emph{admit with full funding}.  For the latter, I also mention funding sources (e.g., RA from prof. X or TA). I also include my comments to support the decision. Finally, I also recommend very strong students for the University Presidential fellowship (\autoref{sec:fellowships}).

% The adcom chair will look all evaluations for each application and makes a decision (likely following the decision of the reviewers).  If there are discrepancies such as one voted reject but another full funding acceptance (such cases are rare, but do happen), the chair might ask the reviewers to coordinate to reach a concensus.

@@

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a field that is not CS but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is highly valued in CS programs in the US, and experiences in these areas can make you stand out and get noticed from reviewers.

@@

\section{PhDs in the US vs. Other Countries}\label{sec:non-us-differences}

@@

\emph{MS Degree requirement}:  as mentioned in \autoref{sec:msrequirement} and \autoref{sec:time}, CS PhD programs in the US do not require MS degree.  In contrast, many other countries do require MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

\emph{Project proposal}: in many countries, you have to choose a project and adviser \emph{during} the application process (e.g., you write a proposal to a potential adviser). In the US, you often start your PhD without an adviser or project and find them later. This is because as someone who just finished your undergraduate study, it is unlikely that you know enough about research to do proposal or picking an adviser.  You will have two initial years to explore and find an adviser and research topic.

\emph{Course work}: in the US you will spend the first couple of years taking classes and explore potential adviser and research topics. In other countries, you (who already have an MS) start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Also, in the US you also have to pass a series of exams, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}. In other countries, you might not have these exams or only have to do a few of them.

\emph{Funding}:  In many countries stipend comes from the university or from the gov't. These funding often have a fixed duration (e.g., 3 or 4 years).  In the US (\autoref{sec:funding}), stipend (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

\emph{Faculty position after PhD}: In other countries, PhD graduates interested academia typically apply for additional research appointments, i.e., postdocs, and then consider faculty position. In the US, PhD graduates often apply directly for faculty position (postdoc for US graduates is no longer a popular option as it was before).

>>>>

\paragraph{Interviews} Just like a job application, you likely will get interviews from schools. The most common case is that a prof. is interested in working with you and wants to chat to make decision, e.g., to offer RA (\autoref{sec:ra})). In some cases, the interview is done by several professors, e.g., to see if a student fits in their group or to recruit a very strong student to their program. The interview can also be done by adcom members to see if the student fit their program.
%The professor might include one of their current students as part of the interview.

@@

\textbf{Advice}: Treat the interview as an informal chat. Have elevator pitches about your research experiences and interests. You might also want have some 5-minute presentation about your research. If a prof. asks you to read a paper, do it and be prepared to discuss it. You can also ask if you need to prepare for a coding problem. Finally, prepare some thoughtful questions to ask, e.g., about the program, the university, or the professor's research.

@@

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only a few would have NSF Graduate (GRFP) or Microsoft fellowships . In fact, these are so prestigious that even if you didn't get it but make it to the final round or even \emph{``honorable mentioning''}, school will still mention you on their website, and you should put it on your CV.

@@

\begin{commentbox}[Hung:]
  I always encourage the students I admitted to talk with my students and the students of other faculty in other schools who admitted them. You will unlikely hear straight-out complaints from current students in a prof’s group. But sometimes what is important are things that they (current students) don’t tell you. Pay attention to their ``level of excitement'' being in the group.
\end{commentbox}

@@

Here is some general advice for finding a prof.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research that prof. and contact them (\autoref{sec:contact}).

@@

Another effective way is taking graduate-level courses in the topics you are interested in.  Profs teaching \emph{special topics courses} and \emph{research seminars} might be recruiting students---giving you even a higher chance. Do well in the class, answer questions, talk to the prof after classes, etc---being stand out.  Many profs, including myself, prefer taking in new students this way.  It gives both the prof and student more time, e.g., a whole semester, to work and evaluate the relationship before making any commitment (sounds a bit like a marriage!).

@@

\paragraph{Not a good fit}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if there is no faculty who is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students already), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

For example, the 2024 admission cycle has a huge surge in students interested in AI and NLP (thanks, ChatGPT!). Consequently, AI/NLP faculty might be overwhelmed and cannot consider many candidates, even those with excellent profiles.

@@

\paragraph{Overqualified or Lack Interest}  This might be paradoxical and is the opposite of aiming too high. However if adcom believes that you are likely to get admitted and go to a better program (i.e., they are just your ``safety'' schools), they might not be so enthusiastic to admit you and want to save the spot for someone else.

@@

%\section{How exactly do you evaluate an application?}\label{sec:ievaluate}

@@

% I first skim the summary for school, GPA, standard tests and look for red flags like low GPA or test scores that do not meet the minimum requirements. Then I scan transcripts for low/failing grades on CS and related courses.  For these I make comments such as ``- unknown international school with good GPA, too many low grades on CS and related courses" or ``+ very good GPA from top school XYZ". I do not look at test details as the summary already gives the main scores.

% Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experience to distinguish between strong LoRs from the rest). My comments for LoRs include ``+ very strong LoRs from well-known prof. XYZ talks about research experiences'' or ``- generic LoRs, mostly about course work''.  I also note any unusual or unique details in the LoRs, e.g., ``+ LoR mentioned about student's significant contribution on well-known open-source project X''. Sometimes a LoR came directly from a GMU professor who mentioned they already know the student and intend to offer RA, which is a very strong signal for admission (and my comment would be ``Prof. X making RA offer'').

% I skim over CV, looking for paper publications, research work experiences, and other relevant achievements such as winning prizes in well-known international programming/math competitions.  My notes for CV include things such as ``+ has X papers at top venues'' or ``+ gold medal at X''. Sometimes I also note about unusual or unique experiences, e.g., ``+ has a patent on X''.

% Next is the SOP, which I read carefully (but if it is bad then it would be a quick read). My notes for SOP include things such as ``- SOP mostly on course work, no clear motivation, generic" , or ``+ very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned publications then I would have ``+ has X papers at top venues" or ``+/- has X papers but at unknown places''.  I will not go online to search for the papers (unless they upload them as sample papers).  I also skim over sampling papers, essentially looking over the abstract to see if reads well, and make notes such as ``+ have a paper under submission to unknown venue, but reads well". If the student mentioned a project, I might also look at it (e.g., if it is on Github) and make notes such as ``+ has project on Github with X stars''.

% At this point I have enough to make up my decision, which I enter into the evaluation system. Typically my decisions are either \emph{reject} or \emph{admit with full funding}.  For the latter, I also mention funding sources (e.g., RA from prof. X or TA). I also include my comments to support the decision. Finally, I also recommend very strong students for the University Presidential fellowship (\autoref{sec:fellowships}).

% The adcom chair will look at all evaluations for each application and make a decision (likely following the decision of the reviewers).  If there are discrepancies such as one voting reject but another full funding acceptance (such cases are rare, but do happen), the chair might ask the reviewers to coordinate to reach a consensus.

@@

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a non-CS field but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS (\autoref{sec:urm})? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is highly valued in CS programs in the US, and experiences in these areas can make you stand out and get noticed from reviewers.

@@

\section{PhD in the US vs. Other Countries}\label{sec:non-us-differences}

@@

\emph{MS requirement and PhD duration}:  as mentioned in \autoref{sec:msrequirement} and \autoref{sec:time}, CS PhD programs in the US do not require MS degree.  In contrast, many other countries require having an MS degree before joining a PhD program.  This means that US PhD programs are longer (5--7 years, 2 of which are coursework) than other countries (3--4 years, no coursework).

\emph{Project proposal}: in many countries, you have to choose a project and adviser \emph{during} the application process (e.g., you write a proposal to a potential adviser). You start your research right from the beginning.

In the US, you often start your PhD without an adviser or project and find them later. You will have two initial years to explore and find an adviser and research topic. And then you often explore projects with that adviser before settling on one (can take a couple of years).

\emph{Course work}: as mentioned above in the US you will spend the first couple of years taking classes and exploring potential adviser and research topics. 
After that, you also have to pass a series of exams during your PhD, e.g., qualifying exam, comprehensive exam, thesis proposal defense\footnote{The word ``ABD'' (all but dissertation) is used in the US to refer to a PhD candidate who have finished all course work and exams and only need to write and defend their dissertation.}.

In other countries, you start your research right away, e.g., you immediately work on the research project you proposed with the adviser you chose. Moreover, you might not have exams like those in the US or only have to do a few of them.

\emph{Funding}:  In many countries funding comes from the university or the gov't. This funding often has a fixed duration (e.g., 3 or 4 years).  In the US (\autoref{sec:funding}), funding (e.g., RA) comes directly from your adviser (no fixed duration).  There are also fewer TA opportunities in European universities compared to the US.

\emph{Academic position after PhD}: In other countries, PhD graduates interested in academia typically apply for additional research appointments, i.e., postdocs, and then consider faculty positions.

In the US, PhD graduates often apply directly for faculty positions (postdoc for US graduates is no longer a popular option as it was before).
